---
title: "Jittered points and summary statstics plots"
author: Kasia Kedzierska
date: 14 May 2020
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

Always, at the begining we should start with loading the packages and setting up the defaults. As you see here, I have my preferences about the theme and legend position. We won't be using the latter, but it's a force of habit for me to add the following lines to every notebook I set up. 

```{r, warning=FALSE, message=FALSE}
# load the precious libraries
library(tidyverse)
library(ggsci)

# set the defauts for plots
theme_set(theme_classic())
theme_update(legend.position = "bottom")
```

Now, let's create a simulated data to recreate the plots.

```{r}
set.seed(7)
df <- 
  data.frame(
    sample_name = c(rep("sample_A", 15), rep("sample_B", 5)),
    gene_1 = c(rnorm(15, 1), rnorm(5, 3)),
    gene_2 = c(rnorm(15, 10, 5), rnorm(5, 10, 2)),
    gene_3 = c(rnorm(15, 10, 1), rnorm(5, 9, 3))
  ) %>%
  pivot_longer(names_to = "gene_id", 
               values_to = "expression", 
               -sample_name)
```

Firstly, let's create a plot with all observations and their means. 

```{r}
# points with mean
df %>%
  ggplot(aes(sample_name, expression, color = gene_id)) +
  geom_point(position = position_jitterdodge(), size = 2) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean, 
               geom = "crossbar", alpha = .7,
               width = 0.5, color = "grey60") +
  facet_wrap(~gene_id) +
  scale_color_nejm() +
  labs(title = "Example jitter with mean",
       y = "Normalised expression") +
  theme(text = element_text(size = 14), title = element_text(size = 18), 
        axis.title.x = element_blank(), legend.position = "none",
        panel.border = element_rect(colour = "black", fill = NA))
```

Now, the plot with the boxplot. 

The default ggplot `geom_boxplot()` will show upper quartile, median and lower quartile in a rectangle. The lines from the upper quartile extend to largest observation less than or equal to upper quartile `+ 1.5 * IQR`. Similarly, for the `ymin` will extend to smallest observation greater than or equal to lower quartile `- 1.5 * IQR`. Full documentation, and exmaples of how to modify the plot, is available [here](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

Additionally, since we are plotting the data under the boxplot, there is no need for us to show the outliers, so we are setting `outlier.shape = NA`.

```{r}
# points with boxplot
df %>%
  ggplot(aes(sample_name, expression, color = gene_id)) +
  geom_point(position = position_jitterdodge(), size = 2) +
  geom_boxplot(width = .2, alpha = .4, outlier.shape = NA, 
               color = "grey60") +
  facet_wrap(~gene_id) +
  scale_color_nejm() +
  labs(title = "Example jitter with boxplot",
       y = "Normalised expression") +
  theme(text = element_text(size = 14), title = element_text(size = 18), 
        axis.title.x = element_blank(), legend.position = "none",
        panel.border = element_rect(colour = "black", fill = NA)) 
```

